---
title: "data exploration"
author: "Rosemary Juarez"
date: 2024-01-29
---
#NOTE:
#bri's code down below 

## Preparation

Set-up chunk

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

Load libraries

```{r}
library(tidyverse)
library(here)
```

Read in Data

```{r}
# read in tree ring data
rwi_long <- read_csv(here("~/../../capstone/climatree/dataverse-archive/rwi_long.csv"))

# read in site summary table
site_summary <- read_table(here("~/../../capstone/climatree/dataverse-archive/site_summary.tab"))

# read in species metadata
species_metadata <- read_table(here("~/../../capstone/climatree/dataverse-archive/species_metadata.tab"))

# essential cwd data
essential_cwd <- read_csv(here("~/../../capstone/climatree/dataverse-archive/essentialcwd_data.csv"))

```

## Selecting Tree Species

```{r, join}

# remove quotes from `collection_id` column in site data ----
site_summary <- site_summary %>% 
  mutate(collection_id = str_remove_all(collection_id, "\""))

# capitalize`sp_id` column in species metadata ----
species_metadata <- species_metadata %>% 
  rename(sp_id = species_id) %>% 
  mutate(sp_id = toupper(sp_id))

# join data ----
rwi_species <- rwi_long %>%
  
  # drop extra columns
  select(-c(rwi_ar, rwi_nb)) %>% 
  
  # join rwi_long to site_summary by collection_id
  left_join(y = site_summary, by = 'collection_id') %>% 
  
  # drop extra columns
  select(collection_id, core_id, tree, core, year, rwi, rwl, sp_id) %>%
  
  # left join to species metadata, by species id
  left_join(y = species_metadata, by = 'sp_id') %>% 
  
  # drop extra columns
  select(-c(source, spp)) %>% 
  
  # remove quotes from strings in genus and species column
  mutate(genus = str_remove_all(genus, "\""),
         species = str_remove_all(species, "\"")) %>% 
  
  # combine genus and species into a single column
  mutate(genus_species = str_c(genus, " ", species))

```

```{r, plot join}
# create df with counts of collection ID
sp_count <- plyr::count(rwi_species$sp_id) %>% 
  rename(sp_id = x)

# plot top 30 collection ID counts
sp_count %>% 
  slice_max(order_by = freq, n = 15) %>% 
  ggplot() +
  geom_col(aes(x = fct_reorder(sp_id, freq), y = freq)) +
  labs(x = "Species ID",
       y = "Number of Samples",
       title = "15 Most Sampled Species") +
  geom_text(aes(x = fct_reorder(sp_id, freq), y = freq,
                label = freq), hjust = 1.2, color = "white") +
  coord_flip() +
  theme_minimal() 
```

**NOTE:** I found a discrepancy in the number of species. There are some NAs in the `species_metadata` that affected the new `genus_species` column I created. Therefore, there's more species ID's than there are associated species and genus names. The plot above is created using only species ID, but if we want to use species as a searchable map item we have to consider how we'll manage NAs. In addition to the NAs, there are some columns where genus and species are listed as "et al".

```{r}
# print differences in the number of unique species
glue::glue("Number of unique genus_species: {length(unique(rwi_species$genus_species))}
           Number of unique sp_id: {length(unique(rwi_species$sp_id))}
           Number of unique common_name:{length(unique(rwi_species$common_name))}")

# view distribution of NAs
colSums(is.na(rwi_species))
```
